Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-12 ind_files/03_field_eqns.tex
Minor phrasing changes, add some references
Minor phrasing changes

If you know and care about enough mathematics for this to really bother you, see~\citep{champagnat_unifying_2006} for a much more rigorous treatment that avoids using informal tools such as functional derivatives and functional equivalents of Fokker-Planck equations in favor of a probabilistic approach grounded in (measure-theoretic) Markov and martingale theory.}. The state space of the stochastic process thus becomes infinite-dimensional, which complicates matters slightly. The principal objects of interest here are \emph{functionals} $F[x, \phi^{(t)}]$ which take in a scalar $x$ representing the trait value of interest, and a function $\phi^{(t)}$ representing the population at time $t$. Thus, whereas in the previous section we were interested in how a function $f(x(t))$ changes based on the change in an input variable $x(t)$ (the population), we are now interested in how a functional $F[\phi^{(t)}]$ changes with the change in an input function $\phi^{(t)}$. The mathematics for these sorts of processes is an active area of research and is comparatively far from well developed. The mathematically rigorous formulation of the kinds of processes we study here falls in the realm of measure-valued branching processes, and is highly technical and rather inaccessible unless one is already comfortable with advanced measure-theoretic notions~\citep{champagnat_unifying_2006,champagnat_individual_2008}. This means that the existing formalism, while admirable in its generality and mathematical rigor, is rather unusable for most biologists, who do not have formal training in analysis (but see~\cite{week_white_2021} for a very friendly introduction to the major ideas through heuristics). One can, however, make progress if they are willing to take some mathematical leaps of faith and sacrifice rigor for the sake of accessibility and heuristic understanding. I adopt this attitude below and hope that all the (rather pedantic) questions of rigor, well-posedness, existence, etc. will be sorted out by some clever mathematicians in the future. Physicists use the term `field' for functions of the form $f(x,t):\mathbb{R}^{n} \times [0,\infty) \to \mathbb{R}^{m}$, where $\mathbb{R}^{n}$ represents space and $[0,\infty)$ represents time. They then call models which describe such functions `field theories'. In physics jargon, the stochastic process we will formulate $\{\phi^{(t)}\}_{t \geq 0}$ when viewed as a sequence of functions $\{\phi^{(t)}(y)\}_{t \geq 0}$ thus describes a (scalar) `stochastic field', and the formalism we will develop below is a `stochastic field theory' of evolution, where physical space has been replaced by an abstract trait space. This is closely related to the area of physics called `statistical field theory', the analog of quantum field theory for systems with a large number of classical particles. Stochastic field theories over physical space have recently been used in biology to model brain function~\citep{bressloff_stochastic_2010} and collective motion~\citep{o_laighleis_minimal_2018}. In the following sections, I will rely heavily on a heuristic object called the functional derivative $\delta F/\delta \phi$. The functional derivative is an \emph{ad hoc}, somewhat informal notion, defined indirectly as the unique object that obeys, for any `nice' function $\rho$

@@

\section{Stochastic trait frequency dynamics in the infinite-dimensional case}
In chapter \ref{chap_BD}, we also derived SDEs for the trait frequency dynamics of the complete stochastic case using It\^o's formula. Doing the same for quantitative traits in our framework is tricky because it requires us to find a version of It\^o's formula that holds for SPDEs of the form \eqref{functional_langevin}. The formulation of It\^o formulas and/or an infinite-dimensional stochastic calculus for general function (measure) valued stochastic processes is an active area of research in pure mathematics~\citep{da_prato_stochastic_2014}, and without further information on the nature of the domain $\mathcal{M}_{K}(\mathcal{T})$, it is not clear (to me at least) whether an It\^o's formula exists for our case in general. If $\mathcal{A}^{\pm}(x | \phi)$ in \eqref{functional_FPE} are Gaussian functions, then it has been proven~\citep{week_white_2021} that the SDEs `work as expected' if we take $m \to \infty$ in \eqref{nD_eqn_for_frequencies} (see \ref{sec_fun_theorems_mean} and \ref{sec_fun_theorems_var} in chapter \ref{chap_unification}). However, carrying out a general derivation is beyond the scope of this thesis.

>>>>

If you know and care about enough mathematics for this to really bother you, see~\citep{champagnat_unifying_2006} for a much more rigorous treatment that avoids using informal tools such as functional derivatives and functional equivalents of Fokker-Planck equations in favor of a probabilistic approach grounded in (measure-theoretic) Markov and martingale theory.}. The state space of the stochastic process thus becomes infinite-dimensional, which complicates matters slightly. The principal objects of interest here are \emph{functionals} $F[x, \phi^{(t)}]$ which take in a scalar $x$ representing the trait value of interest, and a function $\phi^{(t)}$ representing the population at time $t$. Thus, whereas in the previous section we were interested in how a function $f(x(t))$ changes based on the change in an input variable $x(t)$ (the population), we are now interested in how a functional $F[\phi^{(t)}]$ changes with the change in an input function $\phi^{(t)}$. The mathematics for these sorts of processes is an active area of research and is comparatively far from well developed. The mathematically rigorous formulation of the kinds of processes we study here falls in the realm of measure-valued branching processes, and is highly technical and rather inaccessible unless one is already comfortable with advanced measure-theoretic notions~\citep{champagnat_unifying_2006,champagnat_individual_2008}. This means that the existing formalism, while admirable in its generality and mathematical rigor, is rather unusable for most biologists, who do not have formal training in analysis (but see~\cite{week_white_2021} for a very friendly introduction to the major ideas through heuristics). One can, however, make progress if they are willing to take some mathematical leaps of faith and sacrifice rigor for the sake of accessibility and heuristic understanding. I adopt this attitude below and hope that all questions of rigor, well-posedness, existence, etc. will be sorted out by some clever mathematicians in the future. Physicists use the term `field' for functions of the form $f(x,t):\mathbb{R}^{n} \times [0,\infty) \to \mathbb{R}^{m}$, where $\mathbb{R}^{n}$ represents space and $[0,\infty)$ represents time. They then call models which describe such functions `field theories'. In physics jargon, the stochastic process we will formulate $\{\phi^{(t)}\}_{t \geq 0}$ when viewed as a sequence of functions $\{\phi^{(t)}(y)\}_{t \geq 0}$ thus describes a (scalar) `stochastic field', and the formalism we will develop below is a `stochastic field theory' of evolution, where physical space has been replaced by an abstract trait space. This is closely related to the area of physics called `statistical field theory', the analog of quantum field theory for systems with a large number of classical particles. Stochastic field theories over physical space have recently been used in biology to model brain function~\citep{bressloff_stochastic_2010} and collective motion~\citep{o_laighleis_minimal_2018}. In the following sections, I will rely heavily on a heuristic object called the functional derivative $\delta F/\delta \phi$. The functional derivative is an \emph{ad hoc}, somewhat informal notion, defined indirectly as the unique object that obeys, for any `nice' function $\rho$

@@

\section{Stochastic trait frequency dynamics in the infinite-dimensional case}
In chapter \ref{chap_BD}, we also derived SDEs for the trait frequency dynamics of the complete stochastic case using It\^o's formula. Doing the same for quantitative traits in our framework is tricky because it requires us to find a version of It\^o's formula that holds for SPDEs of the form \eqref{functional_langevin}. The formulation of It\^o formulas and/or an infinite-dimensional stochastic calculus for general function (measure) valued stochastic processes is an active area of research in pure mathematics~\citep{da_prato_stochastic_2014}, and without further information on the nature of the domain $\mathcal{M}_{K}(\mathcal{T})$, it is not clear (to me at least) whether an It\^o's formula exists for our case in general. However, it bears noting that the `intuitive' It\^o's formula one would expect does indeed hold true for broad classes of Hilbert space valued processes~\citep{prevot_concise_2007,da_prato_stochastic_2014,liu_stochastic_2015}. If $\mathcal{A}^{\pm}(x | \phi)$ in \eqref{functional_FPE} are Gaussian functions, then it has been proven~\citep{week_white_2021} that the SDEs `work as expected' if we take $m \to \infty$ in \eqref{nD_eqn_for_frequencies} (see section \ref{sec_fun_theorems} in chapter \ref{chap_unification}). However, carrying out a general derivation is beyond the scope of this thesis.
